Solar Flares Across the Spectrum: Observations and Instrumentation

### 2025-04-11 chapter2/chap2_2.tex
Local changes

The remarkable technological progress attained in the last few decades has yielded significant benefits in the form of highly advanced imaging, spectroscopic, and polarimetric instruments designed for astronomical observations. These instruments have empowered us with the ability to scrutinize the Sun with exceptional detail. Specifically, the space based observatories have added new avenues of observing the Sun. The studies within this thesis are performed using multi-wavelength imaging and spectroscopic observations of solar flares recorded from space. This chapter describes the details of the instruments and techniques used for our analysis.

@@

The Solar Dynamics Observatory \citep[SDO;][]{sdo} is a NASA mission designed to understand the causes of solar variability at various spatio-temporal and wavelength scales, as well as its impacts on Earth and the near-Earth environment. It is part of NASA’s Living With a Star program and houses multiple instruments onboard to address its numerous scientific goals. Among the three instruments onboard {\it SDO}, we have extensively used data from two instruments – the Atmospheric Imaging Assembly (AIA) and the Helioseismic and Magnetic Imager (HMI).

@@

Solar Orbiter \citep[SO;][]{so} is a joint mission between the European Space Agency (ESA) and NASA, launched in February 2020. The Spectrometer/Telescope for Imaging X-rays (STIX) \citep[STIX,][]{stix,stix1} is designed to study the Sun's X-ray emissions, providing crucial insights into solar flares and other high-energy phenomena. STIX incorporates a set of collimators and detectors to capture high-resolution images of the Sun in X-rays. These images allow us to pinpoint the locations and intensities of X-ray emissions associated with solar flares. The instrument covers a broad range of X-ray energies, from approximately 4 to 150 keV, probing both thermal and non-thermal X-rays. STIX provides X-ray imaging with a spatial resolution $\sim$ 7 {\arcsec} at the shortest wavelengths using Fourier imaging techniques. Notably, the spatial resolution depends on the visibilities at various energies. STIX can operate at a temporal cadence as high as 0.1 s. The cadence of observation is adjustable depending on the flaring activity. The instrument operates in two data modes: the low-resolution background monitoring mode in quiet conditions and the high-resolution mode activated during solar flares. STIX works together with other instruments onboard the Solar Orbiter, providing complementary measurements to enhance our understanding of the Sun's dynamic behaviour across different wavelengths and energy ranges.

@@

{\it Hinode} \citep{hinode}, meaning "Sunrise" in Japanese, is a space mission launched by the Japan Aerospace Exploration Agency (JAXA) in collaboration with NASA and the UK Space Agency in September 2006. Also known as Solar-B before its launch, {\it Hinode} is dedicated to studying the Sun's magnetic field and its dynamic behaviour, focusing on understanding the mechanisms driving solar activity and space weather phenomena. The X-Ray Telescope \citep[XRT;][]{xrt} is one of the key instruments aboard the {\it Hinode} spacecraft.

@@

Geostationary Operational Environmental Satellites (GOES) are a series of weather satellites operated by the National Oceanic and Atmospheric Administration (NOAA) in the United States. These satellites continuously monitor weather conditions across the Americas, including the United States, Canada, Mexico, and parts of South America. GOES data are used by meteorologists and weather forecasters to monitor and predict weather conditions, track severe weather events, and issue warnings and advisories to the public. The continuous monitoring provided by GOES satellites helps improve the accuracy and timeliness of weather forecasts. In addition to monitoring terrestrial weather, GOES satellites also monitor space weather conditions, including solar flares, coronal mass ejections (CMEs), and geomagnetic storms. This information is crucial for forecasting space weather events and assessing their potential impact on satellite communications, navigation systems, and power grids. Among the numerous instruments on-board {\it GOES}, we have used data from two instruments extensively {--} X-Ray Sensor (XRS) and Solar Ultra Violet Imager (SUVI).

@@

In Eqn.~\ref{eqn:dem}, $\mathrm{g_{i}}$ is the counts observed in some specific filter (usually in units of $\mathrm{DN.s^{-1}.pix^{-1}}$), $\mathrm{\delta g_{i}}$ is the observed error, $\mathrm{R_{i}}$ is the temperature response function of the specific filter and DEM(T) describes the thermal distribution of the local plasma  (usually in units of $\mathrm{cm^{-5}.K^{-1}}$). For multiple filter observations, this problem can be posed as a matrix equation in the form of $\Vec{g}~=~\mathbb{R}~\Vec{\zeta}\implies \Vec{\zeta}~=~\mathbb{R}^{-1}\Vec{g}$. However, because of the ill-posed and under-determined nature of the inverse problem, any attempt to invert this set of equations and recovering the DEM results in a significant increase in the noise.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Doppler Velocity from Spectra}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The Solar Ultraviolet Imaging Telescope}\label{sec:suit}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

The Solar Ultraviolet Imaging Telescope \citep[SUIT;][]{ghosh16,article}, iss one of the seven payloads aboard the Aditya-L1 mission \citep{adityal1, aditya} led by the Indian Space Research Organization (ISRO). Launched on September 2, 2023, the satellite orbits around the Sun-Earth L1 point in a halo orbit. Equipped with eleven science filters, comprising eight narrow-band and three broad-band filters, SUIT possesses a distinctive capability to explore various heights within the solar photosphere and chromosphere. This capability aids in understanding the diverse physical processes involved in the transport of mass and energy across different layers of the Sun. Additionally, SUIT offers a unique opportunity to spatially measure the solar spectral irradiance in the Near Ultraviolet (NUV) range, a critical aspect in advancing our comprehension of Sun-climate relationships.

{\suit} is designed to deliver continuous full-disk and Region of Interest (RoI) solar images, boasting a plate scale of 0.7{\arcsec}. It possesses the ability to track RoIs while compensating for the effects of differential rotation, as outlined by \cite{suit_algo}. Additionally, SUIT is equipped with onboard intelligence to detect and localize flares, and it can autonomously adjust exposure times to prevent saturation. The primary scientific inquiries targeted by SUIT include \citep{suit_science, suit_main}:

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{SUIT \& Solar Flares}\label{sec:suit_and_flare}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

The Solar Ultraviolet Imaging Telescope \citep[SUIT;][]{ghosh16,article} is one of the seven payloads onboard the Aditya-L1 mission \citep{adityal1} of the Indian Space Research Organization (ISRO). With its 11 science filters (3 broadband and eight narrow band), SUIT will have the capability to probe different heights of solar atmosphere in photosphere and chromosphere to help us understand the various processes that transport mass and energy from one layer to another. Through SUIT imaging, we would be able to resolve solar flares spatially on the surface of the Sun, for the first time in near ultra-violet (NUV), which will help us to address the questions regarding their build-up and triggering mechanism. In addition, to measure the spatially resolved solar spectral irradiance within the wavelength range that is central for studying the Chemistry of oxygen and ozone in the Stratosphere of Earth's atmosphere.

It has been shown that the majority of flare energy emerges at the visible and UV wavelength range \citep{woods06}. \cite{woods04} showed that about 77\% of the energy is released in the wavelength range > 200 nm, and only ~ 23\% is seen in extreme ultraviolet (EUV) and soft X-ray (SXR), i.e., below 200 nm \citep{Nei_1989,neidig93,kretzschmar11}. Although the energy content in hard X-ray (HXR) is a tiny fraction of the total energy budget, they are still crucial in understanding the energization process \citep{holeman11}. However, to develop a comprehensive understanding of solar flares, it is mandatory to perform multi-wavelength studies of all kinds of flares. This may have implications on the physics of the origin of solar flares and different physics processes and contribute to the solar spectral irradiance as a function of the solar activity cycle. Although we have been observing Sun and Solar flares in various wavelengths, the spectral energy distribution of the radiated energy from the flares is still very poorly understood. The first solar flares were observed from the ground in the visible domain \citep{carrington1859,neidig93}. It is also well known that the flare emission in the visible domain occurs mainly in $H\alpha$ and Ca \Romannum{2} lines \citep{canfield90,falchi92,heinzel94}. However, the lesser understood component of the visible and Near Ultra Violet (NUV) emission is the enhancement of the continuum. The study of the white-light (WL) flares has proven to be very difficult because they have a very short duration and low contrast against the background, making their observation from Earth rare and of poor quality. Also, the flares in NUV are not observable from any ground-based instrumentation as most of the NUV gets absorbed in the upper atmosphere, thus requires space-based observations.

The origin of the WLFs, i.e., the physical process responsible for generating the continuum and its contribution to the overall energy distribution, is still highly uncertain. The question remains whether WLFs are photospheric phenomena due to $H^{-}$ free-free emission or chromospheric phenomena due to H free-bound emission. The more recent studies further constrain the origin of the WLFs to be Chromospheric phenomena, as it has been shown that the WL and HXR foot point centroids are cospatial. Similarly, \cite{krucker15} constrained the cospatial WL and HXR foot points within the chromosphere for three flares. With the help of SUIT, we would be able to localize the WL flares and resolve them on various parts of the Solar disk, which along with observations from Interface Region Imaging Telescope (IRIS), Helioseismic and Magnetic Imager (HMI), would help us localize the source of the WL foot points and also comment on the formation mechanism of the WL itself.

@@

Finally, one of the major question of interest is how the Solar flares affect the Spectral Solar Irradiance (SSI) and Total Solar Irradiance (TSI) variability from a short to much longer, Solar Cycle timescale. \cite{kretzschmar10,kretzschmar11} performed statistical studies using a large number of flare observations across a wide energy range. For this purpose, they used the full Sun observations of Solar flux from Solar and Heliospheric Observatory (SoHO), three visible Solar irradiances from VIRGO/Solar Photometer (SPM) pass bands centred on 402 nm, 500 nm, and 862 nm, respectively, from 1996 to 2008. Additionally, they also use the EUV irradiance in the ranges 0.1-50 nm and 26-34 nm measured by SOHO/Solar EUV Monitor \citep{judge98} and SXR measurements from GOES satellites. They showed that stacked TSI variation profiles during solar flares show variation at more than 2 sigma level during the peak flare time, indicating the presence of flare signals in the TSI measurements.

@@

This allows us to quantify the solar variability induced be solar flares of various timescales. With the help of SUIT, we would be able to localize the flare locations and study the change in SSI from the local environment in the 11 science filters. This information combined with TSI measurements can allow us to quantify the effect of flares of various energy scales on the TSI variability of the Sun. As, both the TSI and SSI variability directly or indirectly couples with various atmospheric parameters, we can also study the effect the flares have on them. For example, the Earth's atmospheric chemistry and composition respond to any changes in solar UV output in a very nonlinear fashion \citep{haigh07}. Since the number of flares also shows a change with solar activity, it is prudent to ask how much flares contribute to the solar spectral irradiance in NUV. This is particularly important because the irradiance in NUV plays a key role in heating the upper and middle layers of the Earth's atmosphere directly and their coupling with the Stratosphere. It also directly influences the middle and lower atmosphere chemistry and composition via the Ozone-Oxygen cycle.

>>>>

The remarkable technological progress attained in the last decades has yielded significant benefits in the form of highly advanced imaging, spectroscopic, and polarimetric instruments designed for astronomical observations. These instruments have empowered us with the ability to scrutinise the Sun with exceptional detail. Specifically, space-based observatories have added new avenues for observing the Sun. The studies within this thesis are performed using multi-wavelength imaging and spectroscopic observations of solar flares recorded from space. This chapter briefly describes and provides relevant details of instruments whose observations are used in this thesis. Moreover, we also discuss the data analysis techniques used for our analysis.

@@

The solar Dynamics Observatory \citep[SDO;][]{sdo} is a NASA mission designed to understand the causes of solar variability at various spatio-temporal and wavelength scales, as well as its impacts on Earth and the near-Earth environment. It is part of NASA’s Living With a Star program and houses multiple instruments onboard to address its numerous scientific goals. Among the three instruments onboard {\it SDO}, we have extensively used data from two instruments – the Atmospheric Imaging Assembly (AIA) and the Helioseismic and Magnetic Imager (HMI).

@@

Solar Orbiter \citep[SO;][]{so} is a joint mission between the European Space Agency (ESA) and NASA, launched in February 2020. The Spectrometer/Telescope for Imaging X-rays \citep[STIX,][]{stix,stix1} is designed to study the Sun's X-ray emissions, providing crucial insights into solar flares and other high-energy phenomena. STIX incorporates a set of collimators and detectors to capture high-resolution images of the Sun in X-rays. These images allow us to pinpoint the locations and intensities of X-ray emissions associated with solar flares. The instrument covers a broad range of X-ray energies, from approximately 4 to 150 keV, probing both thermal and non-thermal X-rays. STIX provides X-ray imaging with a spatial resolution $\sim$ 7{\arcsec} at the shortest wavelengths using Fourier imaging techniques. Notably, the spatial resolution depends on the visibilities at various energies. STIX can operate at a temporal cadence as high as 0.1 s. The cadence of observation is adjustable depending on the flaring activity. The instrument operates in two data modes: the low-resolution background monitoring mode in quiet conditions and the high-resolution mode activated during solar flares. STIX works together with other instruments onboard the solar Orbiter, providing complementary measurements to enhance our understanding of the Sun's dynamic behaviour across different wavelengths and energy ranges.

@@

{\it Hinode} \citep{hinode}, meaning "Sunrise" in Japanese, is a space mission launched by the Japan Aerospace Exploration Agency (JAXA) in collaboration with NASA and the UK Space Agency in September 2006. Also known as solar-B before its launch, {\it Hinode} is dedicated to studying the Sun's magnetic field and its dynamic behaviour, focusing on understanding the mechanisms driving solar activity and space weather phenomena. The X-Ray Telescope \citep[XRT;][]{xrt} is one of the key instruments aboard the {\it Hinode} spacecraft.

@@

Geostationary Operational Environmental Satellites (GOES) are a series of weather satellites operated by the National Oceanic and Atmospheric Administration (NOAA) in the United States. These satellites continuously monitor weather conditions across the Americas, including the United States, Canada, Mexico, and parts of South America. GOES data are used by meteorologists and weather forecasters to monitor and predict weather conditions, track severe weather events, and issue warnings and advisories to the public. The continuous monitoring provided by GOES satellites helps improve the accuracy and timeliness of weather forecasts. In addition to monitoring terrestrial weather, GOES satellites also monitor space weather conditions, including solar flares, coronal mass ejections (CMEs), and geomagnetic storms. This information is crucial for forecasting space weather events and assessing their potential impact on satellite communications, navigation systems, and power grids. Among the numerous instruments on-board {\it GOES}, we have used data from two instruments extensively {--} X-Ray Sensor (XRS) and solar Ultra Violet Imager (SUVI).

@@

In Eqn.~\ref{eqn:dem}, $\mathrm{g_{i}}$ is the counts observed in some specific filter (usually in units of $\mathrm{DN.s^{-1}.pix^{-1}}$), $\mathrm{\delta g_{i}}$ is the observed error, $\mathrm{R_{i}}$ is the temperature response function of the specific filter and {\bf $\mathrm{\zeta(T)}$} describes the thermal distribution of the local plasma  (usually in units of $\mathrm{cm^{-5}.K^{-1}}$). For multiple filter observations, this problem can be posed as a matrix equation in the form of $\Vec{g}~=~\mathbb{R}~\Vec{\zeta}\implies \Vec{\zeta}~=~\mathbb{R}^{-1}\Vec{g}$. However, because of the ill-posed and under-determined nature of the inverse problem, any attempt to invert this set of equations and recovering the DEM results in a significant increase in the noise.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Doppler and Non-thermal Velocity from Spectra}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The solar Ultraviolet Imaging Telescope (SUIT)}\label{sec:suit}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

The solar Ultraviolet Imaging Telescope \citep[SUIT;][]{ghosh16,article}, is one of the seven payloads aboard the Aditya-L1 mission \citep{adityal1, aditya} led by the Indian Space Research Organization (ISRO). Launched on September 2, 2023, the satellite orbits around the Sun-Earth L1 point in a halo orbit. Equipped with eleven science filters, comprising eight narrow-band and three broad-band filters, SUIT possesses a distinctive capability to explore various heights within the solar photosphere and chromosphere. This capability aids in understanding the diverse physical processes involved in the transport of mass and energy across different layers of the Sun. Additionally, SUIT offers a unique opportunity to spatially measure the solar spectral irradiance in the Near Ultraviolet (NUV) range, a critical aspect in advancing our comprehension of Sun-climate relationships.

{\suit} is designed to deliver continuous full-disk and Region of Interest (RoI) solar images, boasting a plate scale of 0.7{\arcsec}. It possesses the ability to track RoIs while compensating for the effects of differential rotation, as outlined by \cite{suit_algo}. Additionally, SUIT is equipped with onboard intelligence to detect and localize flares, and it can autonomously adjust exposure times to prevent saturation. The primary scientific inquiries targeted by SUIT include \citep{suit_main}:

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{SUIT \& solar Flares}\label{sec:suit_and_flare}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

The solar Ultraviolet Imaging Telescope \citep[SUIT;][]{ghosh16,article} is one of the seven payloads onboard the Aditya-L1 mission \citep{adityal1} of the Indian Space Research Organization (ISRO). With its eleven science filters (three broadband and eight narrow band), SUIT will have the capability to probe different heights of solar atmosphere in photosphere and chromosphere to help us understand the various processes that transport mass and energy from one layer to another. Through SUIT imaging, we would be able to resolve solar flares spatially on the surface of the Sun, for the first time in near ultra-violet (NUV) {\bf 200 {-- 400} nm regime}, which will help us to address the questions regarding their build-up and triggering mechanism. In addition, to measure the spatially resolved solar spectral irradiance within the wavelength range that is central for studying the Chemistry of oxygen and ozone in the Stratosphere of Earth's atmosphere.

It has been shown that the majority of flare energy emerges at the visible and UV wavelength range \citep{woods06}. \cite{woods04} showed that about 77\% of the energy is released in the wavelength range > 200 nm, and only ~ 23\% is seen in extreme ultraviolet (EUV) and soft X-ray (SXR), i.e., below 200 nm \citep{Nei_1989,neidig93,kretzschmar11}. Although the energy content in hard X-ray (HXR) is a tiny fraction of the total energy budget, they are still crucial in understanding the energization process \citep{holeman11}. However, to develop a comprehensive understanding of solar flares, it is mandatory to perform multi-wavelength studies of all kinds of flares. This may have implications on the physics of the origin of solar flares and different physics processes and contribute to the solar spectral irradiance as a function of the solar activity cycle. Although we have been observing Sun and solar flares in various wavelengths, the spectral energy distribution of the radiated energy from the flares is still very poorly understood. The first solar flares were observed from the ground in the visible domain \citep{carrington1859,neidig93}. It is also well known that the flare emission in the visible domain occurs mainly in $H\alpha$ and Ca \Romannum{2} lines \citep{canfield90,falchi92,heinzel94}. However, the lesser understood component of the visible and Near Ultra Violet (NUV) emission is the enhancement of the continuum. The study of the white-light (WL) flares has proven to be very difficult because they have a very short duration and low contrast against the background, making their observation from Earth rare and of poor quality. Also, the flares in NUV are not observable from any ground-based instrumentation as most of the NUV gets absorbed in the upper atmosphere, thus requires space-based observations. {\bf The attenuation is usually given by

%%%%%%%%%
\begin{equation}
    \mathrm{I~=~I_{0}e^{-\alpha d}}
\end{equation}
%%%%%%%%%

where the I and $\mathrm{I_{0}}$ are the initial and transmitted intensity respectively, `d' is the distance traveled by the incident flux. The attenuation coefficient ($\mathrm{\alpha}$) is defined in units of $\mathrm{length^{-1}}$ and given by the Linke-Borne formula \citep{linke22, middleton51}, $\mathrm{\alpha~=~A\lambda^{-B}}$, which depends on various physical parameters of the atmosphere, e.g., density, temperature, pressure, composition of the air column etc. Under average atmospheric conditions, the attenuation coefficient varies significantly in the wavelength range 200 {--} 400 nm, varying from > $\mathrm{4~km^{-1}}$ around $\sim$ 200 nm, to < $\mathrm{0.4~km^{-1}}$ around $\sim$ 400 nm \citep[see Fig.7, Fig.83 and the accompanying discussions in][]{baum_thes}.}

The origin of the WLFs, i.e., the physical process responsible for generating the continuum and its contribution to the overall energy distribution, is still highly uncertain. The question remains whether WLFs are photospheric phenomena due to $H^{-}$ free-free emission or chromospheric phenomena due to H free-bound emission. The more recent studies further constrain the origin of the WLFs to be Chromospheric phenomena, as it has been shown that the WL and HXR foot point centroids are cospatial. Similarly, \cite{krucker15} constrained the cospatial WL and HXR foot points within the chromosphere for three flares. With the help of SUIT, we would be able to localize the WL flares and resolve them on various parts of the solar disk, which along with observations from Interface Region Imaging Telescope (IRIS), Helioseismic and Magnetic Imager (HMI), would help us localize the source of the WL foot points and also comment on the formation mechanism of the WL itself.

@@

Finally, one of the major question of interest is how the solar flares affect the Spectral solar Irradiance (SSI) and Total solar Irradiance (TSI) variability from a short to much longer, solar Cycle timescale. \cite{kretzschmar10,kretzschmar11} performed statistical studies using a large number of flare observations across a wide energy range. For this purpose, they used the full Sun observations of solar flux from solar and Heliospheric Observatory (SoHO), three visible solar irradiances from VIRGO/Solar Photometer (SPM) pass bands centred on 402 nm, 500 nm, and 862 nm, respectively, from 1996 to 2008. Additionally, they also use the EUV irradiance in the ranges 0.1-50 nm and 26-34 nm measured by SOHO/Solar EUV Monitor \citep{judge98} and SXR measurements from GOES satellites. They showed that stacked TSI variation profiles during solar flares show variation at more than 2 sigma level during the peak flare time, indicating the presence of flare signals in the TSI measurements.

@@

This allows us to quantify the solar variability induced be solar flares of various timescales. With the help of SUIT, we would be able to localize the flare locations and study the change in SSI from the local environment in the 11 science filters. This information combined with TSI measurements can allow us to quantify the effect of flares of various energy scales on the TSI variability of the Sun. As, both the TSI and SSI variability directly or indirectly couples with various atmospheric parameters, we can also study the effect the flares have on them. For example, the Earth's atmospheric chemistry and composition respond to any changes in solar UV output in a very nonlinear fashion \citep{haigh07}. Since the number of flares also shows a change with solar activity, it is prudent to ask how much do flares contribute to the solar spectral irradiance in NUV {\bf over solar cycle timescales}. This is particularly important because the irradiance in NUV plays a key role in heating the upper and middle layers of the Earth's atmosphere directly and their coupling with the Stratosphere. It also directly influences the middle and lower atmosphere chemistry and composition via the Ozone-Oxygen cycle. {\bf In certain extreme cases, a single flare can drive noticeable changes in the upper atmospheric conditions, e.g., the X-class flare on 7th September, 2011 induced a large variation in the E-layer conductivity driven by the $\mathrm{Ly-\alpha}$ radiation \citep[see Fig.10 and the accompanying discussion in][]{milligan20}.}
